---
title: "Data management"
output:
  html_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::knit_hooks$set(
  margin1 = function(before, options, envir) {
    if (before) par(mgp = c(1.5, .5, 0), bty = "n", plt = c(.105, .97, .13, .97))
    else NULL
  },
  margin2 = function(before, options, envir) {
    if (before) par(mgp = c(2, .5, 0), bty = "n", plt = c(.105, .97, .13, .97))
    else NULL
  },
  margin3 = function(before, options, envir) {
    if (before) par(mgp = c(1.5, .5, 0), bty = "n", mai = rep(.1, 4))
    else NULL
  }
)

knitr::opts_chunk$set(echo       = TRUE,
                      margin1    = TRUE,
                      fig.retina = 2,
                      fig.align  = "center")

l <- "en_US.UTF-8"
Sys.setenv(LANGAGE = l)
Sys.setlocale(locale = l)
Sys.setlocale("LC_MESSAGES", l)
```

## Preambule

This document prepares the demographic and epidemiological data necessary for subsequent analyses. These scripts need to be correctly run before attempting any simulations or statistical analyses. It uses the following data sets. 

+ the geographical polygons from [GADM](https://gadm.org).
+ the population counts by local government area from [WorldPop](https://www.worldpop.org).
+ the age-specific fertility rates (`ng2_fertility_data.csv`) from [WorldPop](https://www.worldpop.org).
+ the life table estimates at the country level (`ng0_mortality_data.xlsx`) from [UN Department of Economic and Social Affairs](https://population.un.org/wpp/Download/Standard/Mortality/). 

The first three data sets provide information by local government area (LGA) (775 in Nigeria). The age-specific ferility rates (dataset #3) and life tables (dataset #4) need to be manually downloaded from the provided links. 

### Missing data

+ 32 local government areas do not have age-specific fertility rates. Fertility data for these local government areas are excluded from the aggregated state-level fertility estimates. 

## Packages

The needed packages:

```{r message = FALSE}
library(tidyverse)
library(stars)
library(sf)
library(magrittr)
library(readxl)
library(RColorBrewer)
library(parallel)
```

## Links to local data sets

Here we use the following links to local data.

```{r}
DATA_FOLDER <- file.path("local_data")
ng0_mortality_path <- file.path(DATA_FOLDER, "ng0_mortality_data.xlsx")
ng2_fertility_path <- file.path(DATA_FOLDER, "ng2_fertility_data.csv")
```

Change them accordingly if you want to run the script locally on your computer.

## Spatial polygons from GADM

Downloading the polygons (if necessary) from [GADM](https://gadm.org).

```{r}
if (!file.exists(file.path(DATA_FOLDER, "gadm36_NGA_0_sf.rds"))) {
  download.file("https://geodata.ucdavis.edu/gadm/gadm3.6/Rsf/gadm36_NGA_0_sf.rds", 
                file.path(DATA_FOLDER, "gadm36_NGA_0_sf.rds"))
}
if (!file.exists(file.path(DATA_FOLDER, "gadm36_NGA_1_sf.rds"))) {
  download.file("https://geodata.ucdavis.edu/gadm/gadm3.6/Rsf/gadm36_NGA_1_sf.rds", 
                file.path(DATA_FOLDER, "gadm36_NGA_1_sf.rds"))
}
if (!file.exists(file.path(DATA_FOLDER, "gadm36_NGA_2_sf.rds"))) {
  download.file("https://geodata.ucdavis.edu/gadm/gadm3.6/Rsf/gadm36_NGA_2_sf.rds", 
                file.path(DATA_FOLDER, "gadm36_NGA_2_sf.rds"))
}
```

Loading the polygons:

```{r}
raw_ng0 <- readRDS(file.path(DATA_FOLDER, "gadm36_NGA_0_sf.rds")) %>% # country polygon
  transmute(country = str_squish(NAME_0))

raw_ng1 <- readRDS(file.path(DATA_FOLDER, "gadm36_NGA_1_sf.rds")) %>% # state polygons
  transmute(state = str_squish(NAME_1))

raw_ng2 <- readRDS(file.path(DATA_FOLDER, "gadm36_NGA_2_sf.rds")) %>% # LGA polygons
  transmute(state = str_squish(NAME_1), 
            LGA = str_squish(NAME_2))
```

Simplifying the polygons: 

```{r}
simplified_ng0 <- st_simplify(st_make_valid(raw_ng0), dTolerance=1000)
simplified_ng1 <- st_simplify(st_make_valid(raw_ng1), dTolerance=1000)
simplified_ng2 <- st_simplify(st_make_valid(raw_ng2), dTolerance=1000)
```

Saving the polygons: 

```{r eval=FALSE}
saveRDS(simplified_ng0, file.path(DATA_FOLDER, "ng0.rds"))
saveRDS(simplified_ng1, file.path(DATA_FOLDER, "ng1.rds"))
saveRDS(simplified_ng2, file.path(DATA_FOLDER, "ng2.rds"))
```

Reading the polygons: 

```{r}
ng0 <- readRDS(file.path(DATA_FOLDER, "ng0.rds"))
ng1 <- readRDS(file.path(DATA_FOLDER, "ng1.rds"))
ng2 <- readRDS(file.path(DATA_FOLDER, "ng2.rds"))
```

## WorldPop population data

Downloading the population data from [WorldPop](https://www.worldpop.org). Data can also be manually downloaded from this [link](https://hub.worldpop.org/geodata/summary?id=50493). 

```{r eval=FALSE}
get_worldpop_data <- function(gender, agegrp) {
  link <- paste("https://data.worldpop.org/GIS/AgeSex_structures/Global_2000_2020_Constrained_UNadj/2020/NGA//nga", 
                gender, agegrp, "2020_constrained_UNadj.tif", sep="_")
  file_name <- paste("nga", gender, agegrp, "2020.tif", sep="_")
  download.file(link, file.path(DATA_FOLDER, file_name))
}
genders <- c("f", "m")
agegrps <- c("0", "1", as.character(seq(5, 80, 5)))
input <- expand.grid(gender=genders, agegrp=agegrps)
(parallel::mcmapply(get_worldpop_data, as.character(input[["gender"]]), input[["agegrp"]], mc.cores=5))
```

Here, we compute the population size of each local government area (775) for each combination of gender (`m` or `f`) and age group (`0`, `1-4`, `5-9`,...,`80+`) using the polygons from the GADM dataset to set LGA boundaries and the WorldPop dataset for the population sizes (~1 hr):

```{r eval=FALSE}
lga_population <- function(demo_input, this_lga, this_state, nga_pop, ng=raw_ng2) {
  lga_polygon <- filter(ng, state==this_state, LGA==this_lga) %>% 
    st_crop(nga_pop)
  lga_pop_stars <- nga_pop %>%
    st_crop(lga_polygon) %>%
    st_as_stars()
  
  all_file_names <- paste0("nga_", demo_input[, "gender"], "_", demo_input[, "agegrp"], "_2020_constrained_UNadj.tif")
  n <- c()
  for (ii in all_file_names) {
    n <- c(n, sum(lga_pop_stars[[ii]], na.rm=TRUE))
  }
  lga_pop <- demo_input %>% 
    as.data.frame() %>% 
    mutate(n=n, LGA=this_lga, state=this_state)
  return(lga_pop)
}
all_files <- list.files(path=DATA_FOLDER, pattern = "2020_constrained_UNadj.tif")
nga_pop_stars_proxy <- do.call(c, purrr::map(all_files, function(x) read_stars(file.path(DATA_FOLDER, x))))
input <- expand.grid(gender=c("f", "m"), agegrp=c("0", "1", as.character(seq(5, 80, 5))))
regions <- select(ng2, state, LGA) %>% st_drop_geometry() %>% unique()
nga_pop_list <- list()
for (ii in 1:nrow(regions)) {
  print(paste0(ii, ": ", regions[ii, "state"], ", ", regions[ii, "LGA"]))
  nga_pop_list[[ii]] <- lga_population(demo_input=input, 
                                  this_lga=regions[ii, "LGA"], 
                                  this_state=regions[ii, "state"], 
                                  nga_pop=nga_pop_stars_proxy)
}
nga_pop <- do.call(rbind, nga_pop_list)
rownames(nga_pop) <- NULL
saveRDS(nga_pop, file.path(DATA_FOLDER, "ng2_pop.rds"))
```

Reading in population data...

```{r}
ng2_pop_path <- file.path(DATA_FOLDER, "ng2_pop.rds")
ng2_pop <- readRDS(ng2_pop_path)
```

## LGA fertility data

Let's load the age-specific fertility rate for females of child-bearing age (15-44 years) in each local government area. 

```{r}
ng2_fertility_data <- read.csv(ng2_fertility_path) %>% 
  transmute(state=State, LGA=LGA, agegrp=Age_Group, asfr=ASFR) %>%
  mutate(agegrp=gsub("_", "-", agegrp))
```

Let's check how many local government areas do not have fertility data. 

```{r}
missing_lgas <- ng2_pop %>% 
  anti_join(ng2_fertility_data, by=c("state", "LGA")) %>% 
  select(state, LGA) %>% 
  unique()
nrow(missing_lgas)
```
There are 32 local government areas without age-specific fertility data. We will handle this missing data by excluding these LGAs from the calculation of the aggregated state-level age-specific fertility rates. 

Let's add information on the age distribution of the population to the fertility data.

```{r}
ng2_fertility_data <- ng2_pop %>%
  mutate(agegrp=floor(as.numeric(as.character(agegrp))/5)*5) %>%
  group_by(state, LGA, agegrp, gender) %>%
  summarise(n=sum(n), .groups="keep") %>%
  mutate(agegrp=paste(agegrp, agegrp+4, sep="-")) %>%
  spread(key=gender, value=n) %>%
  mutate(`m/f`=m/f, n=m+f) %>%
  select(state, LGA, agegrp, n, `m/f`) %>% 
  as.data.frame() %>% 
  left_join(ng2_fertility_data, by=c("state", "LGA", "agegrp")) %>% 
  mutate(asfr=ifelse(is.na(asfr), 0, asfr)) %>% 
  mutate(asfr=ifelse((state %in% missing_lgas$state) & (LGA %in% missing_lgas$LGA), 
                     NA, asfr))
```

### Crude birth rates

Let's compute the crude birth rate in each local government area. 

```{r}
ng2_cbr_data <- ng2_fertility_data %>%
  mutate(f=n/(`m/f`+1)) %>%
  group_by(state, LGA) %>%
  summarise(cbr=sum(f*asfr)/sum(n)*1000, .groups="keep")
```

Let's compute the crude birth rate in each state. 

```{r}
ng1_cbr_data <- ng2_fertility_data %>% 
  mutate(f=n/(1+`m/f`), m=n/(1+1/`m/f`)) %>% 
  group_by(state, agegrp) %>% 
  summarise(asfr=sum(asfr*f, na.rm=TRUE)/sum(f, na.rm=TRUE), 
            f=sum(f, na.rm=TRUE), m=sum(m, na.rm=TRUE), .groups="keep") %>%
  ungroup() %>% group_by(state) %>%
  summarise(cbr=sum(asfr*f, na.rm=TRUE)/sum(f+m, na.rm=TRUE)*1000, .groups="keep") %>% 
  as.data.frame()
```

Let's also compute the age-specific fertility rate in each state. 

```{r eval=FALSE}
ng1_fertility_data <- ng2_fertility_data %>% 
  mutate(f=n/(1+`m/f`), m=n/(1+1/`m/f`)) %>% 
  group_by(state, agegrp) %>% 
  summarise(asfr=sum(asfr*f, na.rm=TRUE)/sum(f, na.rm=TRUE), 
            f=sum(f, na.rm=TRUE), m=sum(m, na.rm=TRUE), .groups="keep") %>% 
  select(state, agegrp, asfr) %>% 
  filter(agegrp %in% c("15-19", "20-24", "25-29", "30-34", "35-39", "40-44")) %>% 
  spread(key="agegrp", value="asfr") %>% 
  ungroup()
write_csv(ng1_fertility_data, file=file.path(DATA_FOLDER, "ng1_fertility_data.csv"))
```

## Mortality data

In this section, we estimate the age-specific natural mortality rate function for Nigeria. We start by using life tables provided by the [UN Department of Economic and Social Affairs](https://population.un.org/wpp/Download/Standard/Mortality/). Let's read in the data and retrieve the most recent mortality data for Nigeria. 

```{r}
# load the age-specific mortality rate for Nigeria
ng0_mortality_data <- readxl::read_excel(path=ng0_mortality_path, sheet="ESTIMATES", skip=16) %>%
  select(`Region, subregion, country or area *`, `Period`, `Age (x)`, `Age interval (n)`, 
         `Probability of surviving p(x,n)`) %>%
  setNames(c("region", "period", "age", "age_interval", "surv_prob")) %>% 
  filter(region=="Nigeria", period=="2015-2020") %>% 
  head(-1) %>% 
  transmute(region=as.character(region), age=as.numeric(age), 
            age_interval=as.numeric(age_interval), surv_prob=as.numeric(surv_prob)) %>% 
  mutate(cum_surv_prob=cumprod(surv_prob))
```

Next, let's compute the piece-wise constant age-specific mortality rates. We consider three age groups: $(0, 5)$, $(5, 15)$ and $(15, 80)$. These piece-wise constant age-specific mortality rates will be used to estimate the basic reproductive ratio and the who-acquired-infection-from-whom (WAIFW) matrices in the `R0_estimation.Rmd` analyses. 
```{r}
# compute the piecewise age-specific mortality rate function for R0 estimation
agegrps <- c(5, 15, 80)
ng0_mortality_rates <- rep(NA, length(agegrps))
prev_cum_surv_prob <- 1
prev_agegrp <- 0
for (ii in agegrps) {
  agegrp_pos <- which(ng0_mortality_data$age==ii)-1
  cum_surv_prob <- as.numeric(ng0_mortality_data[agegrp_pos, "cum_surv_prob"])
  surv_prob <- cum_surv_prob/prev_cum_surv_prob
  interval <- ii - prev_agegrp
  mu <- (-1/interval)*log(surv_prob)
  ng0_mortality_rates[which(agegrps==ii)] <- mu
  prev_agegrp <- ii
  prev_cum_surv_prob <- cum_surv_prob
}
```

Saving the mortality data for Nigeria. 

```{r}
saveRDS(ng0_mortality_data, file=file.path(DATA_FOLDER, "ng0_mortality_data.rds"))
saveRDS(ng0_mortality_rates, file=file.path(DATA_FOLDER, "ng0_mortality_rates.rds"))
```

## Aggregation of data at national level

In this section, we aggregate data at the national scale. 

### Population data

```{r}
ng0_pop <- ng2_pop %>% 
  mutate(country="Nigeria") %>% 
  group_by(gender, agegrp) %>% 
  summarise(n=sum(n, na.rm=TRUE), .groups="keep") %>% 
  as.data.frame()
```

### Fertility data

```{r}
ng0_fertility_data <- ng2_fertility_data %>% 
  mutate(country="Nigeria", f=n/(`m/f`+1)) %>% 
  mutate(births=f*asfr) %>% 
  group_by(country, agegrp) %>% 
  summarise(n=sum(n, na.rm=TRUE), f=sum(f, na.rm=TRUE), births=sum(births, na.rm=TRUE), .groups="keep") %>% 
  as.data.frame() %>% 
  mutate(`m/f`=n/f-1, asfr=births/f) %>% 
  select(-f, -births)
```

### Save aggregate data

```{r}
saveRDS(ng0_fertility_data, file.path(DATA_FOLDER, "ng0_fertility_data.rds"))
saveRDS(ng0_pop, file.path(DATA_FOLDER, "ng0_pop.rds"))
```
