---
title: "R0 estimation"
output:
  html_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::knit_hooks$set(
  margin1 = function(before, options, envir) {
    if (before) par(mgp = c(1.5, .5, 0), bty = "n", plt = c(.105, .97, .13, .97))
    else NULL
  },
  margin2 = function(before, options, envir) {
    if (before) par(mgp = c(2, .5, 0), bty = "n", plt = c(.105, .97, .13, .97))
    else NULL
  },
  margin3 = function(before, options, envir) {
    if (before) par(mgp = c(1.5, .5, 0), bty = "n", mai = rep(.1, 4))
    else NULL
  }
)

knitr::opts_chunk$set(echo       = TRUE,
                      margin1    = TRUE,
                      fig.retina = 2,
                      fig.align  = "center")
```

## Preambule

This document focuses on estimating the $R_0$ for rubella in each state of Nigeria using age-stratified serological survey data. 

The R Markdown document `data_management.Rmd` must be successfully run prior to this document because it is responsible for downloading and organizing the following data files: `ng2_pop.rds` (population age and sex structure at local government level), `ng2.rds` (spatial polygons at the local government level), `ng1.rds` (spatial polygons at the state level), `ng1_fertility_data.csv` (age-specific fertility rates at the state level), and `ng0_mortality_rates.rds` (piece-wise constant mortality rate function at the national level). The administrative zone designations of the states are provided as a csv file (`ng1_zone_designations.csv`) in the Github repository. The serological survey data at the state level (`ng1_rubella_seroprevalence_data.csv`) are available upon request. 

Details on methodology are provided in the supplementary materials.

## Packages

The needed packages:

```{r message = FALSE}
library(sf)
library(magrittr)
library(tidyverse)
library(RColorBrewer)
library(rstan)
library(scales)
library(lemon)
rstan_options(auto_write = TRUE)
```

## Links to local data sets

Here we use the following links to data.

```{r}
# directories 
DATA_FOLDER <- file.path("local_data")
HELPER_SCRIPTS_FOLDER <- file.path("helper_scripts")
FIGURES_FOLDER <- file.path("figures")

# derived data from `nigeria_data_organization_exploration.Rmd`
ng2_pop_path <- file.path(DATA_FOLDER, "ng2_pop.rds")
ng2_path <- file.path(DATA_FOLDER, "ng2.rds")
ng1_path <- file.path(DATA_FOLDER, "ng1.rds")
ng1_fertility_data_path <- file.path(DATA_FOLDER, "ng1_fertility_data.csv")
ng0_selected_mortality_rates_path <- file.path(DATA_FOLDER, "ng0_mortality_rates.rds")

# local data 
ng1_rub_sero_path <- file.path(DATA_FOLDER, "ng1_rubella_seroprevalence_data.csv")
ng1_zone_mapping_path <- file.path(DATA_FOLDER, "ng1_zone_designations.csv")
```

Change them accordingly if you want to run the script locally on your computer.

## Helper functions

Here we include some helper functions for the estimation procedures. 

```{r}
# computational scripts
source(file.path(HELPER_SCRIPTS_FOLDER, "compute_cum_hazard.R"))
source(file.path(HELPER_SCRIPTS_FOLDER, "compute_prob_susceptible.R"))
source(file.path(HELPER_SCRIPTS_FOLDER, "compute_avg_life_exp.R"))
source(file.path(HELPER_SCRIPTS_FOLDER, "compute_R0.R"))
```

## Data 
### Nigeria Demographic/Geographical Data

Here, we read in some basic demographic data (age-structure/gender split) at the local government level in Nigeria (see `data_management.Rmd` for details on how the data was collected).

```{r}
ng2_pop <- readRDS(ng2_pop_path)
ng1_fertility_data <- read.csv(ng1_fertility_data_path)
ng0_selected_mortality_rates <- readRDS(ng0_selected_mortality_rates_path)
```

Next, let's read in the spatial polygons data (see `data_management.Rmd` for details on how the data was collected).

```{r}
ng1_zone_mapping <- as.data.frame(read.csv(ng1_zone_mapping_path))
ng2 <- readRDS(ng2_path)
ng1 <- readRDS(ng1_path)
```

### Rubella Seroprevalence Data

Here, we read in and format the rubella serological survey data.

```{r}
ng1_rub_sero_data <- read.csv(file=ng1_rub_sero_path) %>%
  transmute(state=State, age=Age+1, positive=Positive, tested=Tested)
```

## State-level estimation 

### Force of infection

First, we estimate the force of infection in each state for the three age groups (i.e. 0-3 years, 3-15 years and 15+ years). We also simultaneously estimate the rate of loss of maternally-derived passive immunity in infants.

```{r}
compute_piecewise_lambda_fit <- function(group_id_data, sero_data, agegrps) {
  sero_data <- sero_data %>% left_join(group_id_data, by="state") %>% arrange(-desc(state), -desc(age))
  N <- nrow(sero_data)
  K <- length(unique(pull(sero_data, state)))
  gs <- sero_data %>% group_by(state) %>% summarise(count=length(state), .groups="keep") %>% pull(count)
  pos <- sero_data$positive
  tested <- sero_data$tested
  age <- sero_data$age
  agegrp_id <- findInterval(age-1, agegrps)
  num_agegrp <- length(agegrps)
  agegrp <- agegrps
  group_id <- pull(sero_data, group_id)[c(1, 1+cumsum(gs[-length(gs)]))]
  num_groups <- length(unique(group_id))
  input_data <- list(N=N, K=K, gs=gs, pos=pos, num_agegrp=num_agegrp, tested=tested, age=age, 
                     agegrp_id=agegrp_id, num_agegrp=num_agegrp, agegrp=agegrp, group_id=group_id, num_groups=num_groups)
  init <- list(list(lambda=rnorm(K*num_agegrp-(K-num_groups), 0.1, 0.01)), 
               list(lambda=rnorm(K*num_agegrp-(K-num_groups), 0.11, 0.01)),
               list(lambda=rnorm(K*num_agegrp-(K-num_groups), 0.12, 0.01)))
  fit <- stan(file=file.path(HELPER_SCRIPTS_FOLDER, "piecewise_lambda_fit.stan"), 
              data = input_data, chains=3, init=init,
              warmup=2000, iter=4000, cores = 3, control = list(adapt_delta = 0.99), seed=100)
  return(fit)
}
agegrps <- c(0, 3, 15)
states <- unique(pull(ng1_rub_sero_data, state))
input <- data.frame(state=states, group_id=seq_along(states)) %>%
  mutate(group_id=ifelse(state %in% c("Taraba", "Adamawa"), group_id[which(state=="Taraba")], group_id)) %>% 
  mutate(group_id=ifelse(state %in% c("Katsina", "Kano"), group_id[which(state=="Katsina")], group_id)) 
group_ids <- pull(input, group_id) %>% unique() %>% sort()
input$group_id <- sapply(input$group_id, function(x) which(group_ids==x))
```

Fitting the model to seroprevalence data...

```{r eval=FALSE}
MCMC_piecewise_lambda_fit <- compute_piecewise_lambda_fit(group_id_data=input, 
                                                          sero_data=ng1_rub_sero_data,
                                                          agegrps=agegrps)
saveRDS(MCMC_piecewise_lambda_fit, file=file.path(DATA_FOLDER, "MCMC_piecewise_lambda_fit.rds"))
```

Loading the results of the fitting...

```{r}
MCMC_piecewise_lambda_fit <- readRDS(file.path(DATA_FOLDER,
                                               "MCMC_piecewise_lambda_fit.rds"))
```

We first have to reformat the force of infection data. 

```{r eval=FALSE}
all_lambda_data <- rstan::extract(MCMC_piecewise_lambda_fit)[["lambda"]]
states <- unique(pull(ng1_rub_sero_data, state))
get_state_lambda <- function(this_state, num_age_grps, group_id_data) {
  state_pos <- which(this_state==sort(states))
  lambda_ids <- c((seq(1, num_age_grps-1, 1)-1)*length(states)+state_pos, 
                  (num_age_grps-1)*length(states)+group_id_data[which(group_id_data$state==this_state), 2])
  
  state_lambda_data <- all_lambda_data[, lambda_ids]
  return(state_lambda_data)
} 
state_lambda_data <- purrr::map(states, function(x) get_state_lambda(this_state=x, 
                                                                     num_age_grps=3, 
                                                                     group_id_data=input))
names(state_lambda_data) <- states
saveRDS(state_lambda_data, file=file.path(DATA_FOLDER,
                                          "MCMC_piecewise_lambda_fit_lambda_estimates.rds"))
```

```{r}
state_lambda_data <- readRDS(file=file.path(DATA_FOLDER,
                                            "MCMC_piecewise_lambda_fit_lambda_estimates.rds"))
```

Let's also retrieve and save the estimated rate of loss of maternally-acquired passive immunity for rubella across Nigeria. 

```{r eval=FALSE}
omegas <- rstan::extract(MCMC_piecewise_lambda_fit)[["delta"]]
saveRDS(omegas, file=file.path(DATA_FOLDER,"MCMC_piecewise_lambda_fit_omegas.rds"))
```

Let's also retrieve the mean estimates for seropositivity in each age group for each state. 

```{r eval=FALSE}
states <- sort(unique(pull(ng1_rub_sero_data, state))) 
all_seropos_estimates <- rstan::extract(MCMC_piecewise_lambda_fit)[["p"]]
seropos_list <- list()
state_pos <- 1
for (this_state in states) {
  ages <- ng1_rub_sero_data %>% filter(state==this_state) %>% pull(age)
  seropos <- colMeans(all_seropos_estimates[, state_pos:(state_pos+length(ages)-1)])
  seropos <- data.frame(age=ages, seropos=seropos)
  seropos_list[[this_state]] <- seropos
  state_pos <- state_pos + length(ages)
}
saveRDS(seropos_list, file=file.path(DATA_FOLDER,
                                     "MCMC_piecewise_lambda_fit_mean_seropositivity_estimates.rds"))
```

### Basic reproductive number

Now, let's estimate $R_0$. First, we need to compute the initial age distribution in each population. 

```{r}
compute_initial_population <- function(pop_data, this_state, agegrps=c(0, 3, 15)) {
  state_pop_data <- pop_data %>% 
    filter(state==this_state) %>% 
    group_by(state, agegrp) %>% 
    summarise(n=sum(n, na.rm=TRUE), .groups="keep") %>% 
    as.data.frame() %>% 
    mutate(agegrp=as.numeric(as.character(agegrp)))
  age5_pop <- state_pop_data[which(state_pop_data$agegrp==1), "n"]
  split_data <- data.frame(state=this_state, 
                           agegrp=1:4, 
                           n=age5_pop/4)
  state_pop_data <- state_pop_data %>% 
    filter(agegrp!=1) %>% 
    rbind(., split_data) %>% 
    arrange(-desc(agegrp)) %>% 
    mutate(interval=findInterval(agegrp, agegrps)) %>% 
    group_by(state, interval) %>%
    summarise(n=sum(n), .groups="keep") %>% 
    mutate(agegrp=agegrps[interval]) %>%
    as.data.frame()
  return(state_pop_data)
}
states <- unique(pull(ng1_rub_sero_data, state))
init_pop_list <- lapply(states, function(x) compute_initial_population(pop_data=ng2_pop, this_state=x)) 
names(init_pop_list) <- states
```

Now, we can estimate the $R_0$ in each state. 

```{r eval=FALSE}
agegrps <- c(0, 3, 15, Inf)
contact_mat <- matrix(c(1, 1, 3, 1, 2, 3, 3, 3, 3), nrow=3, ncol=3, byrow=TRUE)
deltas <- rstan::extract(MCMC_piecewise_lambda_fit)[["delta"]]
states <- unique(pull(ng1_rub_sero_data, state))
f <- function(this_state) {
  print(this_state)
  lambda_data <- state_lambda_data[[this_state]]
  N <- sum(ng2_pop[which(ng2_pop$state==ifelse(this_state=="FCT", "Federal Capital Territory", this_state)), "n"], na.rm=TRUE)
  initN <- init_pop_list[[ifelse(this_state=="FCT", "Federal Capital Territory", this_state)]] %>% 
    arrange(-desc(agegrp)) %>% 
    pull(n)
  mu <- ng0_selected_mortality_rates
  
  L <- compute_avg_life_exp(mu=mu, agegrps=agegrps[-length(agegrps)])
  equN <- sapply(1:(length(agegrps)-1), function(x)
                 compute_avg_life_exp(mu, agegrps=agegrps[-length(agegrps)], interval=c(agegrps[x], agegrps[x+1])))/L*N
  M_mat <- compute_M_mat(mu=mu, contact_mat=contact_mat, agegrps=agegrps)
  number_samples <- 2000
  for (tt in 1:number_samples) {
    beta_estimate <- compute_betas(lambdas=lambda_data[tt, ], mus=mu, delta=deltas[tt], agegrps=agegrps, 
                                   initN=initN, equN=equN, N=N, L=L, D=10/365, contact_mat=contact_mat)
    saveRDS(beta_estimate, file=file.path(DATA_FOLDER, "R0_beta_estimates", "tmp", paste0(this_state, "_beta_estimate", tt, ".rds")))
  }
  state_beta_estimates <- do.call(rbind, lapply(1:number_samples, function(x) readRDS(file.path(DATA_FOLDER, 
                                                                                              "R0_beta_estimates", "tmp", paste0(this_state, "_beta_estimate", x, ".rds")))))
  state_R0_estimates <- sapply(1:nrow(state_beta_estimates), function(x) 
    compute_R0(agegrps=agegrps, initN=initN, equN=equN, N=N, L=L, D=10/365, contact_mat=contact_mat, 
               betas=state_beta_estimates[x, ], M_mat=M_mat))
  state_estimates <- cbind(state_R0_estimates, state_beta_estimates)
  file_name <- paste0(this_state, "_R0_beta_estimates.rds")
  saveRDS(state_estimates, file=file.path(DATA_FOLDER, "R0_beta_estimates", file_name))
  gc()
}
{parallel::mclapply(as.list(states), f, mc.cores=3)}
state_estimates <- lapply(states, function(x) readRDS(file=file.path(DATA_FOLDER, 
                                                                     "R0_beta_estimates", 
                                                                     paste0(x, "_R0_beta_estimates.rds"))))
names(state_estimates) <- states
R0_estimates <- do.call(cbind, lapply(states, function(x) state_estimates[[x]][, 1])) 
colnames(R0_estimates) <- states
beta_estimates_list <- lapply(states, function(x) state_estimates[[x]][, -1])
names(beta_estimates_list) <- as.character(states)
mu <-  ng0_selected_mortality_rates %>% 
  rep(., each=length(states)) %>% 
  matrix(, nrow=length(states), ncol=3) %>% 
  as.data.frame() %>% 
  setNames(c("mu1", "mu2", "mu3")) %>% 
  mutate(state=states) %>% 
  select(state, dplyr::everything())
saveRDS(R0_estimates, file=file.path(DATA_FOLDER,
                                     "MCMC_piecewise_lambda_fit_R0_estimates.rds"))
saveRDS(beta_estimates_list, file=file.path(DATA_FOLDER,
                                            "MCMC_piecewise_lambda_fit_beta_estimates.rds"))
saveRDS(list(agegrps=agegrps, contact_mat=contact_mat), file=file.path(DATA_FOLDER,
                                                                       "MCMC_piecewise_lambda_fit_contact_structure.rds"))
saveRDS(mu, file=file.path(DATA_FOLDER,
                           "MCMC_piecewise_lambda_fit_mu_estimates.rds"))
```

## National-level estimation

Next, we perform the same analysis but at the national level.

### Force of infection

First, we estimate the force of infection in each state for the three age groups (i.e. 0-3 years, 3-15 years and 15+ years). We also simultaneously estimate the rate of loss of materally-derived passive immunity in infants. 

```{r}
agegrps <- c(0, 3, 15)
ng0_sero_data <- ng1_rub_sero_data %>% 
  group_by(age) %>% 
  summarise(positive=sum(positive), tested=sum(tested), .groups="keep") %>% 
  ungroup() %>% 
  arrange(-desc(age)) %>% 
  mutate(state="Nigeria")
N <- nrow(ng0_sero_data)
pos <- ng0_sero_data$positive
tested <- ng0_sero_data$tested
age <- ng0_sero_data$age
num_agegrp <- length(agegrps)
agegrp <- agegrps
input_data <- list(N=N, pos=pos, num_agegrp=num_agegrp, tested=tested, age=age, 
                   num_agegrp=num_agegrp, agegrp=agegrp)
init <- list(list(lambda=rnorm(num_agegrp, 0.1, 0.01)), 
             list(lambda=rnorm(num_agegrp, 0.11, 0.01)),
             list(lambda=rnorm(num_agegrp, 0.12, 0.01)))
```

Fitting the data...
```{r eval=FALSE}
ng0_MCMC_piecewise_lambda_fit <- stan(file=file.path(HELPER_SCRIPTS_FOLDER, "national_piecewise_lambda_fit.stan"), 
                                      data = input_data, chains=3, init=init,
                                      warmup=2000, iter=4000, cores = 3,
                                      control = list(adapt_delta = 0.99), seed=100)
saveRDS(ng0_MCMC_piecewise_lambda_fit, 
        file=file.path(DATA_FOLDER, 
                       "ng0_MCMC_piecewise_lambda_fit.rds"))
```

Loading the fitting data...

```{r}
ng0_MCMC_piecewise_lambda_fit <- readRDS(file=file.path(DATA_FOLDER, 
                                                        "ng0_MCMC_piecewise_lambda_fit.rds"))
```

Let's plot the force of infection function:

```{r}
ng0_lambda_data <- rstan::extract(ng0_MCMC_piecewise_lambda_fit)[["lambda"]]
saveRDS(ng0_lambda_data, file.path(DATA_FOLDER, 
                                   "ng0_MCMC_piecewise_lambda_fit_lambda_estimates.rds"))
```

Let's also retrieve and save the estimated rate of loss of maternally-acquired passive immunity for rubella across Nigeria. 

```{r}
omegas <- rstan::extract(ng0_MCMC_piecewise_lambda_fit)[["delta"]]
saveRDS(omegas, file=file.path(DATA_FOLDER,
                               "ng0_MCMC_piecewise_lambda_fit_omegas.rds"))
```

Let's also retrieve the mean estimates for seropositivity in each age group for each state. 

```{r}
ng0_seropos_estimates <- rstan::extract(ng0_MCMC_piecewise_lambda_fit)[["p"]]
ages <- ng0_sero_data$age
ng0_seropos <- colMeans(ng0_seropos_estimates)
ng0_seropos <- data.frame(age=ages, seropos=ng0_seropos)
saveRDS(ng0_seropos, file=file.path(DATA_FOLDER,
                                    "ng0_MCMC_piecewise_lambda_fit_mean_seropositivity_estimates.rds"))
```

### Basic reproductive number

Now, let's estimate $R_0$. First, we need to compute the initial age distribution in each population. 

```{r}
ng0_init_pop <- do.call(rbind, init_pop_list) %>% 
  as.data.frame() %>% 
  group_by(interval, agegrp) %>% 
  summarise(n=sum(n), .groups="keep") %>% 
  as.data.frame()
```

Now, we can estimate the $R_0$ in each state. 

```{r eval=FALSE}
agegrps <- c(0, 3, 15, Inf)
contact_mat <- matrix(c(1, 1, 3, 1, 2, 3, 3, 3, 3), nrow=3, ncol=3, byrow=TRUE)
deltas <- rstan::extract(ng0_MCMC_piecewise_lambda_fit)[["delta"]]
N <- sum(ng0_init_pop$n)
initN <- ng0_init_pop$n
mu <- ng0_selected_mortality_rates
L <- compute_avg_life_exp(mu=mu, agegrps=agegrps[-length(agegrps)])
equN <- sapply(1:(length(agegrps)-1), function(x)
  compute_avg_life_exp(mu, agegrps=agegrps[-length(agegrps)], interval=c(agegrps[x], agegrps[x+1])))/L*N
M_mat <- compute_M_mat(mu=mu, contact_mat=contact_mat, agegrps=agegrps)
beta_estimates_list <- list()
estimate_beta <- function(tt) {
  beta_estimate <- compute_betas(lambdas=ng0_lambda_data[tt, ], mus=mu, 
                                 delta=deltas[tt], agegrps=agegrps, 
                                 initN=initN, equN=equN, N=N, L=L, D=10/365, 
                                 contact_mat=contact_mat)
  saveRDS(beta_estimate, file=file.path(DATA_FOLDER, "R0_beta_estimates", 
                                        "tmp", paste0("Nigeria", "_beta_estimate", tt, ".rds")))
}
number_samples <- 2000
{parallel::mclapply(as.list(1:number_samples), estimate_beta, mc.cores=3)}
ng0_beta_estimates <- do.call(rbind, lapply(1:number_samples, function(x) 
  readRDS(file.path(DATA_FOLDER, "R0_beta_estimates", "tmp", 
                    paste0("Nigeria", "_beta_estimate", x, ".rds"))))) %>% 
  as.data.frame() %>% 
  setNames(c("beta1", "beta2", "beta3"))
ng0_R0_estimates <- sapply(1:nrow(ng0_beta_estimates), function(x) 
  compute_R0(agegrps=agegrps, initN=initN, equN=equN, N=N, L=L, D=10/365, contact_mat=contact_mat, 
             betas=as.numeric(ng0_beta_estimates[x, ]), M_mat=M_mat))

saveRDS(ng0_R0_estimates,
        file=file.path(DATA_FOLDER,
                       "ng0_MCMC_piecewise_lambda_fit_R0_estimates.rds"))
saveRDS(ng0_beta_estimates,
        file=file.path(DATA_FOLDER,
                       "ng0_MCMC_piecewise_lambda_fit_beta_estimates.rds"))
```